\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-10-11 paper/main.tex

  Reproducibility is a core requirement of modern scientific research.
  For computational research, reproducibility means that code should
  produce the same results, even when run on different systems.  A
  standard approach to ensuring reproducibility entails packaging a
  project's dependencies along with its primary code base.  Existing
  solutions vary in how deeply these dependencies are specified,
  ranging from virtual environments (which specify all Python package
  versions), to containers (which also specify the operating system),
  to virtual machines (which also specify hardware layers of the
  system).  Each of these existing solutions requires installing or
  setting up a system for running the desired code that must be
  packaged alongside the primary code base.  Here we propose a
  lighter-weight solution than virtual environments: the
  \texttt{davos} library.  When used in combination with a
  notebook-based Python project, the \texttt{davos} library provides a
  mechanism for specifying (and automatically installing) the correct
  package versions of the project's dependencies.  This enables researchers to
  share a complete reproducible environment using a single Jupyter
  notebook file.

@@

Code sharing is a core component of the open science movement that has
inspired the development of a full ecosystem of tools and packages.
However, sharing code, in and of itself, does not guarantee that
others will be able to \textit{reproduce} the desired results.  For
example, research code often requires installing other software
packages that extend the implementation language's basic
functionality.  Within the Python community~\citep{vanR95}, external
packages that are published in the most popular
repositories~\citep{pypi, Anac12} are associated with version numbers
and tags that enable users to guarantee that they are installing
exactly the same code across different computing environments.
Despite that it is \textit{possible} to manually install the intended
version numbers of every dependency of a Python script or package,
doing so may cause conflicts within the user's computing environment
that interfere with the functionality of \textit{other} code.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text \textbf{Python
    scripts} (.py files) provide the
  most basic ``system'' for sharing raw code.  Scripts may reference
  external libraries, but those libraries are not automatically
  installed on other users' systems, nor is any version checking
  performed by default.  \textbf{Jupyter notebooks} (.ipynb files) comprise embedded text,
  executable code, and media (including rendered figures, code output,
  etc.).  When the \textbf{\texttt{davos} library} is imported into a Jupyter
  notebook, the notebook's functionality is extended to automatically
  install the required external libraries (at their correct versions,
  when specified).  \textbf{Virtual environments} install an isolated copy of
  Python and all required dependencies.  This typically requires
  defining a \texttt{requirements.txt} file that lists all
  dependencies (including version numbers) along with an environment
  (.yml) file that specifies how the virtual environment should be
  configured.  \textbf{Containers} provide a means of defining an isolated
  environment that includes a complete operating system (independent
  of the user's operating system), in addition to a virtual
  environment or other configurations needed to provide the necessary
  computing environment.  Containers are typically defined using
  specification files (e.g., a plain-text \texttt{dockerfile}) that
  instruct the virtualization engine regarding how to build the
  virtual environment.  \textbf{Virtual machines} provide a complete
  hardware-level simulation of the computing environment.  In addition
  to simulating specific hardware, virtual machines (typically
  specified using binary images files) must also define operating
  system-level properties of the computing environment.  Systems to
  the left of the blue vertical line entail sharing individual
  files, with no additional installation or configuration needed to
  run the target software.  Systems to the right of the red vertical
  line provide high reproducibility by supporting precise control over
  dependencies and versioning.  Notebooks enhanced using the
  \texttt{davos} library are easily shareable and require minimal
  setup costs, while also facilitating high reproducibility by
  enabling precise control over project dependencies.}
\label{fig:code-sharing}
\end{figure}

To facilitate code sharing, the Python community has developed a broad
set of approaches and tools (Fig.~\ref{fig:code-sharing}).  At one
extreme, simply publishing a set of Python scripts (.py files) may
enable others to use or gain insights into the relevant work.  Because
Python is installed by default on most modern operating systems, for
some projects this may be sufficient.  Another popular approach
entails creating JSON files, called Jupyter
notebooks~\citep{KluyEtal16}, that comprise a mix of text, executable
code, and embedded media.  Notebooks may call or import external
scripts or libraries in order to provide a more compact and readable
experience for users.  Each of these systems (Python scripts and
notebooks) provides a convenient means of sharing code, with the
caveat that they do not specify the computing environment in which the
code is executed.  Therefore the functionality of code shared using
these systems cannot be guaranteed across different computing
environments.

At another extreme, virtual machines~\citep{Gold74, AltiEtal05,
  Rose99} provide a hardware-level simulation of the desired system.
Virtual machines are typically isolated from the user's system, such
that installing or running software on a virtual machine does not
impact the user's primary operating system or computing environment.
Containers~\citep[e.g.,][]{Merk14, KurtEtal17} provide a similar
``isolated'' experience.  Although containerized environments do not
specify hardware-level operations, they are typically packaged with a
complete operating system, in addition to a complete copy of Python
and any relevant package dependencies.  Virtual
environments~\citep[e.g.,][]{vanREtal14} also provide a computing
environment that is largely separated from the user's main
environment.  They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproduce an operating system for the runtime environment.  Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees-- at the hardware level, operating
system level, and Python environment level, respectively) that the
relevant code will run similarly for different users.  However, each
of these systems also relies on additional software that can be
resource intensive or burdensome to install or configure.

@@

\subsection{Software architecture}
The \texttt{davos} package consists of two interdependent subpackages. The first, \texttt{davos.core}, comprises a set of modules that implement\comment{provide?} the bulk of the package's core functionality, including pipelines for installing and validating packages, custom parsers for the \texttt{smuggle} statement (see Section \ref{subsec:smuggle}) and onion comment (see Section \ref{subsec:onion}), and a runtime interface for configuring \texttt{davos}'s behavior (see Section \ref{subsec:config}). However, certain critical\comment{other important} aspects of this functionality require (often substantially) different implementation approaches depending on various properties of the notebook environment in which \texttt{davos} is used (e.g., whether the frontend is provided by Jupyter or Google Colaboratory, or which version of IPython \cite{PereGran07} is used by the notebook kernel). To deal with this, environment-dependent parts of core features and behaviors are isolated and abstracted to ``helper functions'' in the \texttt{davos.implementations} subpackage. This second subpackage defines multiple, interchangeable versions of each helper function, organized into\comment{per-environment} modules by the conditions that trigger their use. At runtime, \texttt{davos} detects various features in the notebook environment and selectively imports a single version of each helper function into the top-level \texttt{davos.implementations} namespace, allowing \texttt{davos.core} modules to access the correct implementations for the current notebook environment from a single, constant location. An additional benefit of this design pattern is that it makes adding support for new or updated notebook variants to \texttt{davos} in the future relatively easy.

\comment{
- js_functions.py?
- how parser is registered and deregistered?
- stub files?
- test suite?
- packaged with new PEP \_\_\_ standard (pyproject.toml + setup.cfg; no setup.py)?
}
% Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.

\subsection{Software functionalities}

\subsubsection{The \texttt{smuggle} statement}\label{subsec:smuggle}
Importing \texttt{davos} in a Jupyter notebook enables an additional Python keyword: ``\texttt{smuggle}''.
The \texttt{smuggle} statement can be used as a drop-in replacement for Python's built-in \texttt{import} statement to load libraries, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is not installed locally, \texttt{smuggle} statements can handle missing packages on the fly.
If a smuggled package does not exist in the local environment, \texttt{davos} will install it, expose its contents to Python's \texttt{import} machinery, and load it into the namespace for immediate use.

\subsubsection{The onion comment}\label{subsec:onion}
For greater control over the behavior of \texttt{smuggle} statements, \texttt{davos} defines an additional construct called the ``onion comment''. An onion comment is a special type of inline comment that may be placed on a line containing a \texttt{smuggle} statement to customize how \texttt{davos} searches for the smuggled package locally and, if necessary, downloads and installs it. Onion comments follow a simple syntax based on the ``type comment'' syntax introduced in PEP 484 \cite{vanREtal14}, and are designed to make managing packages with \texttt{davos} intuitive and familiar. To construct an onion comment, simply provide the name of the installer program (e.g., \texttt{pip}) and the same arguments one would use to manually install the package as desired via the command line\comment{(see Fig.~\ref{fig:snippet1})} (see Fig.~\ref{fig:snippet1}, lines \_\_\_-\_\_\_ for examples).
%For examples of \texttt{smuggle} statements and accompanying onion comments, see Figure \ref{fig:snippet1}.
%Figure \ref{fig:snippet1} provides a number of examples of \texttt{smuggle} statements accompanied by

Onion comments are useful when smuggling a package whose distribution name (i.e., the name used when installing it) is different from its top-level module name (i.e., the name used when importing it; e.g., Fig.~\ref{fig:snippet1}, lines \_\_\_-\_\_\_). However, the most powerful use of the onion comment is making \texttt{smuggle} statements \textit{version-sensitive}. If an onion comment includes a version specifier \cite{CoghStuf13} (e.g., Fig.~\ref{fig:snippet1}, lines \_\_\_-\_\_\_), \texttt{davos} will ensure that the version of the package loaded into the notebook matches the specific version requested, or satisfies the given version constraints. If the smuggled package exists locally, \texttt{davos} will extract its version info from its metadata and compare it to the specifier provided. If the two are incompatible (or no local installation is found), \texttt{davos} will install and load a suitable version of the package instead. Onion comments can similarly be used to smuggle specific VCS references (e.g., Git \cite{TorvHama05}  branches, commits, tags, etc.; Fig.~\ref{fig:snippet1}, lines \_\_\_-\_\_\_).

%If an onion comment includes a version specifier (\cite{CoghStuf13}; e.g., Fig.~\ref{fig:snippet1}, lines \_\_\_-\_\_\_), \texttt{davos} will first check whether the smuggled package exists locally, and if so, extract its version info from its metadata to compare against the requested version. If the installed version doesn't satisfy the given constraint (or if no local version is found), \texttt{davos} will install and load a suitable version instead. This guarantees the notebook is always run with a specific (or constrained) version of a particular package. Similarly, \texttt{davos}
%If the package exists locally, \texttt{davos} will extract its version from metadata before loading it to determine its version, and if it does not match the requested version (or if no local installation is found),
%If an onion comment includes a version specifier \cite{CoghStuf13}, \texttt{davos} will check whether the smuggled package is installed locally (as usual), and if so, additionally parse its metadata to determine
%Adding a version specifier \cite{CoghStuf13} to an onion comment guarantees the version loaded into the notebook will matches the version specified

\texttt{davos} processes onion comments internally before forwarding arguments to the installer program. In addition to preventing onion comments from being used as a vehicle for shell injection attacks, this allows \texttt{davos} take certain logical actions when particular arguments are passed (e.g., Fig.~\ref{fig:snippet1}, lines \_\_\_-\_\_\_). For example, \texttt{--force-reinstall}, \texttt{-I}/\texttt{--ignore-installed}, and \texttt{-U}/\texttt{--upgrade} will all cause \texttt{davos} to skip searching for a smuggled package locally before installing a new copy; \texttt{--no-input} will disable \texttt{davos}'s input prompts in addition to \texttt{pip}'s; and installing a package into \texttt{<dir>} with \texttt{--target <dir>} will cause \texttt{dir} to be prepended to \texttt{sys.path}, if necessary, so the package can be imported.

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{snippets/snippet1.pdf}
\caption{\small Example \texttt{smuggle} statements and accompanying onion comments.}
\label{fig:snippet1}
\end{figure}

@@

\subsubsection{The \texttt{davos} config}\label{subsec:config}

>>>>

  Reproducibility is a core requirement of modern scientific research.
  For computational research, reproducibility means that code should
  produce the same results, even when run on different systems.  A
  standard approach to ensuring reproducibility entails packaging a
  project's dependencies along with its primary code base.  Existing
  solutions vary in how deeply these dependencies are specified,
  ranging from virtual environments, to containers, to virtual
  machines.  Each of these existing solutions requires installing or
  setting up a system for running the desired code that must be
  packaged alongside the primary code base.  Here we propose a
  lighter-weight solution than virtual environments: the
  \texttt{davos} library.  When used in combination with a
  notebook-based Python project, the \texttt{davos} library provides a
  mechanism for specifying (and automatically installing) the correct
  package versions of the project's dependencies.  The \texttt{davos}
  library also ensures that those versions are in use any time the
  notebook's code is executed.  This enables researchers to share a complete
  reproducible copy of their code within a single Jupyter notebook file.

@@

The same computer code may not behave identically under different
circumstances.  For example, when code depends on external libraries,
different versions of those libraries may function differently.  Or
when CPU or GPU instruction sets differ across machines, the same
high-level code may be compiled into different machine instructions.
Because executing identical code does not guarantee identical
outcomes, code sharing in and of itself is often insufficient for
enabling researchers to reproduce each others' work.

Within the Python~\citep{vanR95} community, external
packages that are published in the most popular
repositories~\citep{pypi, Anac12} are associated with version numbers
and tags that enable users to guarantee that they are installing
exactly the same code across different computing environments.
Despite that it is \textit{possible} to manually install the intended
version numbers of every dependency of a Python script or package, manually
tracking down those dependencies can impose a substantial burden on
the user.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text \textbf{Python
    scripts} (\texttt{.py} files) provide the most basic ``system''
  for sharing raw code.  Scripts may reference external libraries, but
  those libraries must be manually installed on other users' systems.
  Further, any checking needed to verify that the correct versions of
  those libraries were installed must also be performed manually.
  \textbf{Jupyter notebooks} (\texttt{.ipynb} files) comprise embedded
  text, executable code, and media (including rendered figures, code
  output, etc.).  When the \textbf{\texttt{davos} library} is imported
  into a Jupyter notebook, the notebook's functionality is extended to
  automatically install the required external libraries (at their
  correct versions, when specified).  \textbf{Virtual environments}
  install an isolated copy of Python and all required dependencies.
  This typically requires defining a \texttt{requirements.txt} file or
  an environment (\texttt{.yml}) file that that specifies all project
  dependencies (including version numbers of external libraries).
  \textbf{Containers} provide a means of defining an isolated
  environment that includes a complete operating system (independent
  of the user's operating system), in addition to (optionally)
  specifying a virtual environment or other configurations needed to
  provide the necessary computing environment.  Containers are
  typically defined using specification files (e.g., a plain-text
  \texttt{Dockerfile}) that instruct the virtualization engine
  regarding how to build the virtual environment.  \textbf{Virtual
    machines} provide a complete hardware-level simulation of the
  computing environment.  In addition to simulating specific hardware,
  virtual machines (typically specified using binary images files)
  must also define operating system-level properties of the computing
  environment.  Systems to the left of the blue vertical line entail
  sharing individual files, with no additional installation or
  configuration needed to run the target code.  Systems to the right
  of the red vertical line support precise control over dependencies
  and versioning.  Notebooks enhanced using the \texttt{davos} library
  are easily shareable and require minimal setup costs, while also
  facilitating high reproducibility by enabling precise control over
  project dependencies.}
\label{fig:code-sharing}
\end{figure}

Researchers, programmers, and others have developed a broad
set of approaches and tools to facilitate code sharing and
reproducible outcomes (Fig.~\ref{fig:code-sharing}).  At one
extreme, simply publishing a set of Python scripts (.py files) may
enable others to use or gain insights into the relevant work.  Because
Python is installed by default on most modern operating systems, for
some projects this may be sufficient.  Another popular approach
entails creating JSON files, called Jupyter
notebooks~\citep{KluyEtal16}, that comprise a mix of text, executable
code, and embedded media.  Notebooks may call or import external
scripts or libraries in order to provide a more compact and readable
experience for users.  Each of these systems (Python scripts and
notebooks) provides a convenient means of sharing code, with the
caveat that they do not specify the computing environment in which the
code is executed.  Therefore the functionality of code shared using
these systems cannot be guaranteed across different computing
environments.

At another extreme, virtual machines~\citep{Gold74, AltiEtal05,
  vmware} provide a hardware-level simulation of the desired system.
Virtual machines are typically isolated from the user's system, such
that installing or running software on a virtual machine does not
impact the user's primary operating system or computing environment.
Containers~\citep[e.g.,][]{Merk14, KurtEtal17} provide a similar
``isolated'' experience.  Although containerized environments do not
specify hardware-level operations, they are typically packaged with a
complete operating system, in addition to a complete copy of Python
and any relevant package dependencies.  Virtual
environments~\citep[e.g.,][]{vanREtal14} also provide a computing
environment that is largely separated from the user's main
environment.  They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproduce an operating system for the runtime environment.  Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees-- at the hardware level, operating
system level, and Python environment level, respectively) that the
relevant code will run similarly for different users.  However, each
of these systems also relies on additional software that can be
resource intensive or burdensome to install or configure.

@@

\subsection{Software architecture}
The \texttt{davos} package is structured as two sub-packages: a set of ``core" modules that implement...

% Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.

\subsection{Software functionalities}%\label{sec:functionalities}

\subsubsection{The \texttt{smuggle} statement}
Importing \texttt{davos}\comment{in a Jupyter notebook} enables an additional Python keyword: ``\texttt{smuggle}".
The \texttt{smuggle} statement can be used as a drop-in replacement for Python's built-in \texttt{import} statement to load libraries, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is not installed locally, \texttt{smuggle} statements can handle missing packages on the fly.
If a smuggled package does not exist in the local environment, \texttt{davos} will install it, expose its contents to Python's \texttt{import} machinery, and load it into the namespace for immediate use.

\subsubsection{The onion comment}
For greater control over the behavior of \texttt{smuggle} statements, \texttt{davos} defines an additional construct called the \textit{onion comment}. An onion comment is a special type of inline comment that may be placed on a line containing a \texttt{smuggle} statement to customize how \texttt{davos} searches for the smuggled package locally and, if necessary, how it should be installed. Onion comments follow a simple syntax based on the ``type comment" syntax introduced in PEP 484 \cite{vanREtal14} and are designed to make managing packages via \texttt{davos} intuitive and familiar. To construct an onion comment, simply provide the name of the installer program (e.g., \texttt{pip}) and the same arguments one would use to install the package as desired manually via the command line (see Fig.~\ref{fig:snippet1}).

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{snippets/snippet1.pdf}
\caption{\small \textbf{FILL THIS IN...}}
\label{fig:snippet1}
\end{figure}

@@

\subsubsection{The \texttt{davos} config}
